HL-LHC Computing Review:\\Common Tools and Community Software

### 2020-04-30 LHCC/2020/2020-01/ctcs.tex
Break sources into an input file per section

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Common and community software packages, such as ROOT, Geant4 and event
generators have been a key part of the LHC's success so far and
continued development and optimisation will be critical in the future
\cite{stewart_graeme_a_2018_2413005, Ellis:2691414, Albrecht2019}. The
challenges are driven by an ambitious physics programme, notably the LHC
accelerator upgrade to high-luminosity, HL-LHC, and the corresponding
detector upgrades of ATLAS and CMS. The General Purpose Detectors
describe their specific challenges elsewhere; in this document we
address the issues for software that is used in multiple experiments
(usually even more widely than ATLAS and CMS) and maintained by teams of
developers who are either not linked to a particular experiment or who
contribute to common software within the context of their experiment
activity. We also give space to general considerations for future
software and projects that tackle upcoming challenges, no matter who
writes it, which is an area where community convergence on best practice
is extremely useful.

ATLAS and CMS will undergo major detector upgrades and will also
increase their trigger rates for HL-LHC by about a factor of 10; event
complexity will rise, with peak pile-up of 200, far higher than in
Run-2. This places an enormous burden on storage and processing
resources. Current CMOS microprocessor technology is clock speed limited
(due to the failure of Dennard scaling) and, while processors
per-integrated circuit still keeps rising, Moore's Law is expected to
stall during the 2020s. More importantly, the effective runtime related
improvements in computing from CPU servers at sites is likely to be only
about 10\% per year, making the shortfall in processing resources more
severe. As manufacturers struggle to deliver ever-more effective
computing through CPUs the drive to different architectures intensifies,
with GPUs becoming commonplace and increasing interest in even more
specialised architectures, such as TPUs, IPUs and developments that make
FPGA devices more user friendly \cite{acm10.1145/3282307}.
These pose huge challenges for our
community as the programming models and APIs vary widely here and
possible lock-in to a particular vendor's devices is a significant
problem for code sustainability and preservation. Huge work has been
done already to adapt to this changing landscape, e.g. multi-threading
software and GPU codes have been used in production by some experiments
for years~\cite{calafiura2018hep, albrecht2018hepexec}.
In other areas migration is still ongoing. In others it is
not even yet started. Generic heterogeneous programming models have
existed for some time, and new ones are arising, but there is, as yet,
no clear winner, in-part because the performance obtained can be highly
application-dependent. HEP itself will not drive the success of one
model or another, so even if the community converged on a preferred
model, its long term success would not be assured. The C++ standard
itself is likely to lag for many years behind what is required by us in
the area of heterogeneous or even distributed computing. Further,
experiment frameworks (and with knock-on effects to systems like
workload management) will need to adapt to sites that provide
heterogeneous resources. How to do this, and make effective use of
different heterogeneous resources across different sites, remains far
from settled.

For storage systems (see the DOMA and WLCG documents) the pressure on
software is to store as much physics information in as few bytes as
possible, but also to be able to read at very high rates to deliver data
from modern storage technologies into the processing hardware. This
requires critical developments in the storage formats and libraries used
in HEP, e.g. developments like RNTuple for ROOT I/O is likely to be of
great importance for the community \cite{ROOT-2020-HL-LHC}. The likelihood of finding an
off-the-shelf compact and efficient storage format for HEP data is
remote, so investment in smart software to support our PB sized science
data is simply cost effective. Particularly for analysis, which is
usually I/O bound, we have an end-to-end problem from storage
technology, through the software layers, to processing resources that
may well span multiple nodes. Other workflows, which are less dependent
on I/O rates will, nevertheless, have to be adapted to using remote data
where the I/O layer must optimise data transfers and hide latency, e.g.
taking advantage of xrootd's single column streaming ability \cite{xrootd}.

In this increasingly complex environment in which to write software,
there are important problems where sharing information at the community
level is far more efficient. Providing a high level of support in the
build environment for developers, sharing knowledge about how to
measure, and then improve, performance (especially on multiple different
architectures) and sharing best practice for code development can have a
large integrated benefit \cite{couturier2017hep}. This requires improved training and the
development of a curriculum for all developer levels. In the field, such
initiatives are often warmly received, but real support is needed for
those who can put this into practice, also ensuring that their work in
training contributes to their career development and a long term future
in the field \cite{foundation2018hep}. HEP software stacks are already deep and wide and building
these consistently and coherently is also an area where knowledge can be
shared. Support is needed for multiple architectural targets and
ensuring the correct runtime on heterogeneous environments.

This brings up the important question of validation and the need to
improve the security of physics results, which is even more complicated
on heterogeneous platforms, when exact binary compatibility often cannot
be assured. Currently almost every experiment and project has its own
infrastructure for this.

Once software is built, it needs to be shipped worldwide so that the
production workflows can run. CernVM-FS was a huge leap forward for
software distribution and has even been widely adopted outside HEP 
\cite{Blomer_2011, 7310920}.
However, new challenges arise, with container based payloads, scaling
issues and disconnected supercomputer sites. So maintenance and
development needs to be undertaken to support and extend this key
supporting infrastructure for software.

Finally, over the multi-decade lifetimes of HEP experiments, we need to
preserve both the core and analysis software so that results can be
confirmed and updated as the field moves on. There are many exciting
developments based around CernVM-FS
\cite{10.1007/978-3-319-67630-2_52, serverless-cvmfs}, containers and things like analysis
description languages \cite{Ref10}, but these are not yet at the stage of being
settled nor integrated into our day-to-day workflows.

In the rest of this document the main issues associated with the key
parts of the software workflow in high-energy physics are presented,
focusing on those that dominate current resource consumption: physics
event generation, detector simulation, reconstruction and analysis.

\hypertarget{physics-event-generators}{%
\section{Physics Event Generators}\label{physics-event-generators}}

\hypertarget{introduction-1}{%
\subsection{Introduction}\label{introduction-1}}

Physics event generators are essential in HEP. All of the LHC scientific
results, both precision measurements or searches for new physics, depend
significantly on the comparison of experimental measurements to
theoretical predictions computed using generator software.

Using Monte Carlo (MC) techniques, generators allow both the generation
of unweighted events for experimental studies of differential
distributions and the prediction of total cross sections. The
large-scale event generation campaigns of the LHC experiments have
significant computational costs, mainly in terms of CPU resources. The
limited size of simulated samples is a source of major uncertainty in
many analyses and is therefore a limiting factor on the potential
physics output of the LHC programme. This situation will get
significantly worse in the high-luminosity running phase of the LHC
(HL-LHC). The fraction of the CPU resources available to the LHC
experiments which is used for event generation today is currently
estimated to be approximately 5-15\%. As is the case for the other big
consumers of CPU (detector simulation and reconstruction, that have
until now been the main focus), speedups in generator software are also
needed to address the overall resource problem expected at the HL-LHC,
also because more accurate predictions, requiring more complex
calculations (e.g. beyond NLO or with higher jet multiplicities) will be
needed from generators. Many other issues, both technical and
non-technical (e.g. funding, training, careers for those working in this
area), exist~\cite{Alves:2017she,Gen18}.

A HSF Working Group (WG) on generators~\cite{Gen19} was set up at the
beginning of 2019. The main focus of the WG so far has been on gaining a
better understanding of the current situation, and identifying and
prioritising the areas where computing costs can be reduced. In
particular, the WG has been active in analysing the ATLAS and CMS
compute budgets in detail, in profiling MG5\_aMC~\cite{Alw14}, 
Sherpa~\cite{Bot19} and Powheg~\cite{Fri07}, in discussing the possible sharing
of common parton-level samples by ATLAS and CMS, and in reviewing and
supporting the efforts for porting generators to modern architectures
(e.g., MG5\_aMC to GPUs). This last activity is particularly important,
as it has become increasingly clear that being able to run
compute-intensive WLCG software workloads on GPUs would allow the
exploitation of modern GPU-based supercomputers at High Performance
Computing (HPC) centers, and generators look like a natural candidate
for this.

This section gives an overview of the many technical and non-technical
challenges in the generator area and of the work that can be done to
address them. This is finally condensed into a list of a few
high-priority items, for the next 18 months before the LHCC review that
is currently scheduled for fall 2021. A more detailed version of this
contribution, including a more complete list of references, will soon be
uploaded on arxiv and submitted for publication~\cite{Gen20}.

\hypertarget{collaboration-challenges}{%
\subsection{Collaborative challenges}\label{collaboration-challenges}}

1. {The landscape of generator software is extremely varied}. Different
generators are used for different processes. Generating a sample also
involves choices of precision (e.g. Leading Order, LO, or
Next-to-Leading-Order, NLO), hadronization and Parton Shower (PS)
models, underlying event tunes, prescriptions for matching/merging and
simulating particle decays, and other input parameters, chiefly among
them the parton density functions, for which different interfaces exist.
Various combinations of software libraries are thus possible, often
written by different authors and some dating back many years. For a
given process, the LHC experiments often use different software packages
and settings from each other, and a single experiment can generate
events using more than one choice. Many different packages and
configurations may therefore need to be worked on to get cumulative CPU
cost reductions. The large number of external packages also complicates
their long-term maintenance and integration in the experiments software
and workflows, sometimes leading to job failures and computing
inefficiencies. Other packages are also absolutely critical for the
whole generator community and must be maintained, even if their CPU cost
is relatively low (LHAPDF, Rivet, HepMC, etc.).

2. {A very diverse combination of skills/profiles are needed}: theorists
(who create fundamental physics models, and design, develop and optimize
most generator code), experimentalists working in research (who request
types/sizes of samples) and in computing (who implement, monitor and
account execution of workflows on computing resources), software
engineers and system performance experts. This is a richness and
opportunity, as some technical problems are best addressed by people
with specific skills, but it also poses some challenges:

- {Training challenges}\textbf{.} Theorists and experimentalists often
lack formal training in software development and optimization. Software
engineers and experimentalists are often not experts in the theoretical
physics models implemented in MC codes.

- {Communication challenges}. It is difficult to find a shared
terminology and set of concepts to understand one another: notions and
practices that are taken for granted in one domain may be obscure for
others. As an example, there are many articles about the physics in
generators, but software engineers need papers describing the software
modules and overall data/control flow.

- {Career challenges}. Those working in the development, optimization or
execution of generator software provide essential contributions to the
success of the (HL-)LHC physics programme and it is critical that they
get the right recognition and motivation. However, theorists get
recognition on published papers, and may not be motivated to work on
software optimisations that are not "theoretical" enough to advance
their careers. Generator support tasks in the experiments may also not
be enough to secure jobs/funding to experimentalists pursuing a career
in research.

- {Mismatch in usage patterns and optimisation focus}. The way
generators are built and used by their authors is often different from
the way in which they are deployed and integrated by the experiments in
their software frameworks and computing infrastructure.The goals and
metrics of software optimisation work may also differ. Theorists are
mainly interested in calculating cross sections and focus on minimising
the phase space integration time for a given statistical precision. The
LHC experiments run large scale productions of unweighted event
generation, and mainly need to maximize the throughput of events
generated per unit time on a given node.

- {Programming languages}: Attracting collaborators with a computer
science background to work on generators, especially students, may also
be complicated by the fact that critical components of some generator
packages are written in Fortran, which is rarely used in industry and
less popular among developers than other programming languages. Some of
the generators also do not use industry standard version control
systems, making it harder to contribute code.

\hypertarget{technical-challenges-software-and-computing}{%
\subsection{Technical challenges (software and
computing)}\label{technical-challenges-software-and-computing}}

The event generation workflow presents several challenges and
opportunities for improvement.

1. There are many sources of inefficiency in unweighted event
generation:

- {Phase space sampling inefficiency}. Efficient sampling is the most
critical ingredient for efficient unweighted event generation. Many
generic algorithms exist (e.g. VEGAS~\cite{Lep80}, BASES / SPRING~\cite{Kaw86}, 
MINT~\cite{Nas07}, FOAM~\cite{Jad03}), as well as others
developed specifically for a given generator (e.g. MadEvent~\cite{Mal03},
itself based on a modified version of VEGAS, in MG5\_aMC, or COMIX~\cite{Gle08} 
in Sherpa). In general, the larger the dimensionality of the
phase space, the lower the unweighting efficiency that can be achieved:
in W+jets, for instance, the Sherpa efficiency is 30\% for W+0 jets and
0.08\% for W+3jets~\cite{Gao20}. This is an area where research is very
active, and should be actively encouraged, as significant cost
reductions in WLCG compute budgets could be achieved. Improvements in
this area can only start from physics-motivated approaches based on the
knowledge of phase space peaks, but they can be complemented by
brute-force machine learning (ML) algorithmic methods~\cite{Ben17,Bot20,Gao20,Kli18}.

- {Merging inefficiency}. Merging prescriptions (e.g. MLM~\cite{Man02},
CKKW-L~\cite{Lon02} at LO and FxFx~\cite{Fre12}, MEPS@NLO~\cite{Hoe14} at
NLO) imply the rejection of some events, to avoid double counting
between events produced with n\textsubscript{jets}+1 matrix elements and
with n\textsubscript{jets} MEs plus parton showers. The resulting
inefficiencies can be relatively high depending on the process, but they
are unavoidable in the algorithmic strategy used by the underlying
physics modeling~\cite{Alw08}. However, a method like shower-kt MLM can
reduce the merging inefficiency of MLM~\cite{Alw09}.

- {Filtering inefficiency}. An additional large source of inefficiency
is due to the way the experiments simulate some processes, where they
generate large inclusive event samples, which are then ﬁltered on
ﬁnal-state criteria to decide which events are passed on to detector
simulation and reconstruction (e.g. CMS simulations of specific
$\Lambda_{B}$ decays have a 0.01\% efficiency and ATLAS B-hadron
filtering has \textasciitilde10\% efficiency for V+jets). This
inefficiency could be reduced by developing ﬁltering tools within the
generators themselves, designed for compatibility with the requirements
of the experiments. Filtering is an area where LHCb has a lot of
experience and already obtained significant speedups through various
techniques. The speed of color reconnection algorithms is a limiting
factor for simulating rare hadron decays in LHCb.

- {Sample sharing and reweighting}. In addition to removing
inefficiencies, other ways could be explored to make maximal use of the
CPU spent for generation by reusing samples for more than one purpose.
Sharing parton-level samples between ATLAS and CMS is being discussed
for some physics analyses. Event reweighting is already commonly used
(e.g. for new physics searches and some systematic uncertainty
calculations) and could be explored further, though this may require
more theoretical work for samples involving merging or NLO matching~\cite{Mat16}.

- {Negative weights}. Matching prescriptions, like MC@NLO, are required
in (N)NLO calculations to avoid double counting between (N)NLO matrix
elements and parton showers, leading to the appearance of events with
negative weights. This causes a large inefficiency, as larger event
samples must be generated and passed through the experiment simulation,
reconstruction and analysis codes, increasing the compute and storage
requirements. For a fraction r of negative weight events, the number of
events to generate increases by a factor 1/(1-2r)\^{}2: for instance,
with r=25\% (which may be regarded as a worst-case scenario in top quark
pair production~\cite{Fre20}), one needs to generate 4 times as many
events. Negative weights can instead be almost completely avoided, by
design, in another popular matching prescription, POWHEG, which however
is only available for a limited number of processes and describes the
relevant physics to a different degree of precision than MC@NLO 
(see~\cite{Fre20} for a more in-depth discussion). Progress in this area can
only be achieved by theorists: for instance, a new approach for
MC@NLO-type matching with reduced negative weights has recently been
proposed~\cite{Fre20}. It should be noted that negative weights can also
happen at LO because of not-positive-definite parton density function
sets and interference terms, e.g. in effective field theory
calculations.

2. While progress has been made to better understand which areas of
generator software have the highest computational cost, more detailed
accounting of the experiment workloads and profiling of the main
generator software packages would help to further refine R\&D
priorities:

- {Accounting of CPU budgets for generators in ATLAS/CMS}. Thanks to a
lot of effort from the generator teams in both experiments, a lot of
insight into the settings used to support each experiment's physics
programme was gained within the WG, and it is now clear that the
fraction of CPU that ATLAS spends for event generation is somewhat
higher than that in CMS. More detailed analysis of the different
strategies is ongoing. A practical issue is that these figures had to be
harvested from logs and production system databases, which was
particularly difficult for CMS, requiring significant person hours to
extract the required information. It is important to establish better
mechanisms to collect this information, to allow for an easy comparison
between different experiments.

- {Profiling of the generator software setups used for production}.
Another area where the WG has been active is the definition and
profiling of standard generator setups, reproducing those used in
production. Detailed profiling could also be useful to assess the CPU
cost of external parton distribution function libraries~\cite{Kon20}, or
the memory requirements of the software (which may motivate a move to
multithreading or multiprocessing approaches).

3. More generally, as is the case in other areas of HEP, some R\&D on
generator software would certainly be needed to modernise it and make it
more efficient, or even port it to more modern computing 
architectures~\cite{Bau13,Alves:2017she}:

- {Data parallelism, GPUs and vectorization}. The data flow of an MC
generator, where the same (matrix element) function is computed
repeatedly at many phase space points, lends itself naturally to the
data parallel approaches found in CPU vectorised code and in GPU compute
kernels. Porting and optimising generators on GPUs is essential to be
able to use modern GPU-based HPCs (such as SUMMIT, where 95\% of the
compute capacity comes from GPUs). The work done in this direction in
the past on MG5\_aMC, that never reached production quality, is now
being revamped by the WG, in collaboration with the MG5\_aMC team, and
represents one of the main R\&D priorities of the WG. This work is
presently focusing on Nvidia CUDA, but abstraction libraries will also
be investigated. GPUs may also be relevant to ML-based sampling
algorithms and to the pseudo-random number generation libraries used in
all MC generators.

- {Task parallelism, multithreading and multiprocessing}. Generators are
generally executed as single threaded software units. In most cases,
this is not a problem, as the memory footprint of unweighted event
generation is small and usually fits within the 2 GB per core available
on WLCG nodes. However, there are cases where memory is higher than 2 GB
(e.g. DY+jets using Sherpa); this leads to inefficiencies as some
processor cores remain unused, which could be avoided using
multithreading approaches. The fact that some generators are not even
thread safe, may also be a problem, for instance to embed them in
multi-threaded event processing frameworks, such as that of CMS.
Multi-processing approaches may also be useful to speed up the
integration and optimisation step for complex high-dimensional final
states. In particular, a lot of work has been done to implement
MPI-based multi-processing workflows for generators in recent years. For
instance, the scaling of LO-based generation of merged many-jet samples
has been successfully tested and benchmarked on HPC architectures using
both Alpgen~\cite{Chi17} and Sherpa~\cite{Hoe19}; in the latter case, new
event formats based on HDF5, replacing LHEF, have also been instrumental
in the success of these tests. MPI integration has also been completed
for MG5\_aMC~\cite{Mat18}. It should also be noted that, even if HPCs
offer extremely high-speed inter-node connectivity, it is perfectly ok
for WLCG workflows to use these systems as clusters of unrelated nodes.

- {Generic code optimizations}. A speedup of generators may also be
achievable by more generic optimizations, not involving concurrency. It
should be studied, for instance, if data caching~\cite{Kon20} or
different compilers and build strategies may lead to any improvements.

\hypertarget{physics-challenges-e.g.-nnlo}{%
\subsection{Physics challenges (e.g.
NNLO)}\label{physics-challenges-e.g.-nnlo}}

In addition to software issues, important physics questions should also
be addressed about more accurate theoretical predictions (above all NNLO
QCD calculations, but also electroweak corrections) and their potential
impact on the computational cost of event generators at HL-LHC. Some
specific NNLO calculations are already available and used today by the
LHC experiments in their data analysis. With a view to HL-LHC, however,
some open questions remain to be answered, in particular:

1. {NNLO: status of theoretical physics research}. The first question is
for which processes NNLO precision will be available at the time of the
HL-LHC. For example, when would NNLO be expected for 2-\textgreater3 or
even higher multiplicity final states? And for lower multiplicity final
states, where differential NNLO predictions exist but the generation of
unweighted NNLO+PS events is not yet possible. It is important to
clarify the theoretical and more practical challenges in these
calculations, and the corresponding computational strategies and impact
on CPU time needs.

2. {NNLO: experimental requirements for data analysis at HL-LHC}. The
second question is for which final states unweighted event generation
with NNLO precision would actually be required, and how many events
would be needed. One should also ask if reweighting LO event samples to
NNLO would not be an acceptable cheaper alternative to address the
experimental needs.

3. {Size of unweighted event samples required for experimental analysis
at HL-LHC}. Another question, unrelated to NNLO, is in which regions of
phase space the number of unweighted events must be strictly
proportional to the luminosity. For example, in the low pT regions of W
boson production it is probably impossible to keep up with the data, due
to the huge cross section. Alternative techniques should be
investigated, to avoid the generation of huge event samples.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\hypertarget{detector-simulation}{%
\section{Detector Simulation}\label{detector-simulation}}

\hypertarget{introduction-2}{%
\subsection{Introduction}\label{introduction-2}}

In HEP, data analysis, theory model evaluation and detector design
choices rely on detailed detector simulation. Since the start of the
first run the LHC experiments have produced, reconstructed, stored,
transferred and analysed hundreds of billions of simulated events. This
effort required a major fraction of the WLCG computing 
resources~\cite{ALICE-TDR-12, ATLAS-TDR-17, CMS-TDR-7, LHCb-TDR-11, WLCG-CM-Update, ALICE-TDR-019, 
LHCb-TDR-018}

The increase in delivered luminosity in future LHC runs, in particular
with the HL-LHC, will create unique research opportunities by collecting
an order of magnitude more data. At the same time, the demand for
detector simulation will grow accordingly, in order to keep the
statistical uncertainties as low as possible. However, the expected
computing resources will not suffice if current detector simulation is
to be used; the availability of simulated samples of sufficient size
will soon become a major limiting factor as far as new physics
discoveries, for example through precision measurements, are concerned.
Development of faster simulation, therefore, is of crucial importance
and different techniques need to be explored under the assumption that
they will provide a sufficient gain in time performance with a
negligible or acceptable loss in physics accuracy.

The Geant4 simulation toolkit has been the de facto standard for HEP
experiment simulation for physics studies over the last two decades~\cite{Geant4-2003, Geant4-2006, Geant4-2016}.
Designed in the late 1990s to cope with the simulation challenges of the
LHC era, Geant4 introduced flexible physics modeling, exploiting layers
of virtuality. This fostered progress in physics by comparison of
competing models on same energy ranges, and allowed complementary models
to be combined to cover large energy ranges. The simulation applications
of the LHC experiments make use of this toolkit and, as such, most of
the LHC physics program relies on it. It has delivered a physics
precision that allows the experiments to perform precision analysis of
the collected data and to produce new results pushing the boundaries of
our understanding in HEP. At the same time, the code of the toolkit is
becoming old, with some parts written almost thirty years ago, making it
more and more difficult to run efficiently on modern computing
architectures. Any improvement in the critical elements of Geant4 can be
leveraged by the applications that use it. 
Improvements of physics models need to continue to take place to prevent
systematic uncertainties from the simulation from becoming a dominant
factor, however, their exact influence on the precision of physics
analysis is not always easy to quantify. On the other hand, the overall
CPU performance of the current code will not improve drastically just by
relying on modern hardware or better compilers.

Taking all those elements into account, three main development paths have emerged to be
undertaken simultaneously: these have started to be pursued, however,
effort in any one of them is far from sufficient to exploit their
individual potential.

On one hand, it is needed to continue to modernise
and refactor the simulation code by using more optimal, modern
programming techniques. Those improvements, requiring considerable
effort but no major paradigm changes, can reasonably bring several tens
of percent of speed-up, by improving the throughput on modern CPUs,
avoiding constant churn in data and instruction caches. This will
certainly help to produce more simulated events, but it will not be
sufficient to meet the high luminosity needs of the LHC experiments.

On the other hand, different fast simulation techniques, where tracking of
individual particles in given detectors is replaced by some sort of
parameterisation, have already been extensively used for LHC simulation
campaigns, where some precision is traded to achieve higher event
throughput (thus lowering the cost of samples overall). Techniques to
re-use part of the simulated events multiple times have also emerged.
New technologies (like Machine Learning), should certainly be
investigated as fast simulation approaches. All of these will become
more and more significant and will need to be seamlessly integrated with
the detailed simulation provided by Geant4.

Finally, the use of compute
accelerators (like GPUs or FPGAa) is the third emerging path that needs
to be undertaken and that requires intense R\&D effort. We will discuss
more in detail those three development axes in the following sections.

In this section we will outline the developments identified in and since the
Community White Paper~\cite{Albrecht2019, SimFoundation2018hep} focusing
in particular in these three area.

\hypertarget{geant4-rd}{%
\subsection{Geant4 R\&D}\label{geant4-rd}}

The Geant4 particle transport toolkit contains advanced models for
electromagnetic and hadronic physics processes, as well as the
geometrical description of the detectors. The toolkit design principles
were to allow flexibility in future extensions of its own code, as well
as choices for particular applications. It is clear that this
flexibility comes at the price of a performance penalty. It was observed that
writing more compact and more specialised code can lead in certain
places to up to several tens percent speed-up~\cite{GeantV}. The
technical improvements to explore here consist of avoiding too many
virtual layers and improving the instruction and data locality, leading
to better cache use. Going even further as far as specialisation is
concerned, studies suggest that implementing compact, self-contained
transport libraries with a minimal set of physics models (for instance
only electromagnetic interaction for selected particles) with predefined
scoring (to simulate, for instance, the response of an EM calorimeter)
can also bring speed-ups to the complete application~\cite{GeantV}. As
far as architectural modifications are concerned, the design of the
Geant4 track-status machinery prevents the introduction of fine-grained
track parallelism. It is still not clear whether this kind of
parallelism can bring any substantial speed-up, due to the overheads;
however, making the Geant4 kernel and physics processes stateless would
certainly provide the necessary starting points for further R\&D in that
direction. Most of these investigations have started and it is hoped to
see some first results on a year time-scale. More effort, however, will
be required to integrate any eventual changes in the production code in
a timely manner. The GeantV Vector prototype has allowed the assessment
of the achievable speedup using a novel, vectorized approach to particle
transport. The demonstrator of a full electromagnetic shower in
realistic (CMS) geometry has been implemented and the comparison to a
similar Geant4 application has been performed. The conclusion of that
work showed that the main factors in the speedup seem to include better
cache use and tighter code, while the vectorisation impact was much
smaller than hoped for. This unimpressive impact is due to a set of
factors, including the fraction of the current algorithms that could be
vectorised, unresolved challenges for the gather/scatter and tail
handling (to keep the number of events in flight within bound) needed
for large number of shapes, and of course Amdahl's Law applied to
vectorisation where some bottleneck (for example geometry navigation)
could not be tackled without a major additional long term effort~\cite{GeantV}.
At the same time libraries developed in the context of the
GeantV R\&D, e.g. VecGeom, have been successfully integrated in Geant4
and ROOT benefitting the whole HEP software community~\cite{CMS01}. The
cost-over-benefit of more vectorisation of the code has not been deemed
worth the necessary investment also due to the speculation that a
complete rewrite of the physics base could become necessary to fully
exploit it. As a result investing in the optimisation and modernisation
of the Geant4 code has gained even more relevance.

\hypertarget{experiments-applications-and-optimisation-of-the-use-of-geant4}{%
\subsection{Experiments applications and optimisation of the use of
Geant4}\label{experiments-applications-and-optimisation-of-the-use-of-geant4}}

All LHC experiments have dedicated simulation frameworks making use of
the Geant4 toolkit for modelling their detectors and the physics
processes occuring in the experimental setups. Every experiment
configures and uses what is provided by Geant4 for their specific needs.
Effort has been spent by each experiment since the publication of the
Community White Paper~\cite{Albrecht2019, SimFoundation2018hep} to benchmark their 
simulation code and to maximally
exploit the options provided by the Geant4 toolkit to optimise the
performance of their applications~\cite{ATLAS-G4OPT}. All of the various handles provided
are being continuously explored and adopted when deemed useful: range
and physics cuts have been reviewed and customised by all experiments,
shower libraries adopted as baseline in the forward region by ATLAS and
CMS, stepping granularity optimized, magnetic field caching
investigated, neutron russian roulette used for dedicated simulations.
All of them can have major impacts on simulation time and sharing
knowledge and experience between experiments can be of high benefit even
if each experiment has to find its optimal solution.

ATLAS, CMS and LHCb have integrated the Geant4 event-based
multi-threading features within their own multi-threaded frameworks,
harmonizing their different architectural choices with the Geant4
implementation~\cite{CMS02, LHCb01, ATLAS01}, and have or are planning to
use them in their production environment to gain access to a larger set
of distributed resources. ATLAS has been running their multi-process
simulation framework on HPCs systems for production in the
last years~\cite{ATLAS02, Benjamin:2696330} 
while CMS has pioneered the use of their multi-threaded
simulation on available resources including HPCs~\cite{CMS03}. Both
approaches have allowed to exploit a higher number of computing
resources by gaining access to systems with limited memory.

\hypertarget{fast-simulations}{%
\subsection{Fast Simulations}\label{fast-simulations}}

The so-called `fast simulation' techniques consist of replacing the
tracking of individual particles through (a part of) the detector,
including all the physics processes they would undergo, by a
parameterization where the detector response is produced directly as a
function of the incoming particle type and energy. An example of such a
parameterization was implemented many years ago in the context of the H1
experiment~\cite{H1Gflash}. This parameterization, available within the
Geant4 toolkit under the name of the GFLASH library, became the starting
idea for several `custom' implementations, specific for other
calorimeters. Those were interfaced to experiments' simulation
applications through the available `hook' in the Geant4 toolkit. The LHC
experiments implemented, for example, dedicated parametrized response
libraries for some of their calorimeters. Recently, an effort has been
invested in Geant4 to generalize the parameterization formulae, and to
implement automatic procedures of tuning the parameters for specific
detector geometries. This kind of `classical' fast simulation, which
describes the particle shower shape using some complex mathematical
functions with several parameters, will remain an important tool;
however, it is also clear that their overall precision for the different
energy values, especially for highly granular and complex calorimeters
will always be relatively low. The recent developments of Deep
Learning-based techniques have opened an exciting possibility of
replacing those `classical' parameterisations by trained neural networks
that would reproduce the detector response. This approach consists of
training generative models such as Generative Adversarial Networks
(GAN), Variational Auto-Encoders (VAE) or Autoregressive Generative
Networks on the `images' of particle 
showers~\cite{ML001-GAN, aishik_ghosh_2019_3599705, GAN-ATLAS}. The energy
deposition values in the calorimeter cells are considered as `pixels' of
the images that the network is supposed to reproduce. The first studies
and prototypes have shown very promising results, however, the
generalization of the developed models (for different calorimeters, with
different particles incident at different angles with different
energies), still requires further effort. Detailed physics validation of
those tools and understanding of the capability to reproduce the
fluctuations is a prerequisite towards `production-quality' fast
simulation libraries. This work is already ongoing, but will require,
over the next few years, more effort to be invested combining the
physics and Machine Learning expertise. Given that the training of the
deep learning algorithms usually requires using sufficiently large MC
samples produced using standard techniques, the development of the novel
tools does not necessarily remove the need to speed up Geant4.

It is worth pointing out that different fast simulation techniques have
already been extensively used for LHC simulation campaigns. Effort has
been spent in the experiments frameworks to combine as much as possible
fast and detailed implementation. In particular ATLAS and CMS are using
in their baseline production shower libraries for the modelling of
calorimeters in the forward region in combination with the detailed
simulation of the rest of the sub-detectors. In LHCb the re-use of part
of the underlying events for specific signals of interest has been
established in production of samples when appropriate, with particular
care paid to ensure keeping under control bias on statistical
uncertainties~\cite{LHCb02}, its applicability in ATLAS and CMS could be
explored for specific cases. Similarly, the re-use of simulated or real
events to mimic the background to hard-scatter events is also exploited,
where additional challenges exist concerning storage and I/O due to the
handling of the large minimum bias samples needed to model the
additional interactions.

All LHC experiments have explored the use of multi-objective regression
and generative adversarial networks (GANs)~\cite{GAN-ATLAS, GAN-LHCb}, they are now in the process of investigating the use of
fast simulations for other types of detectors than calorimeters, e.g.
Cerenkov based systems~\cite{Lamarr} and Time Projection Chambers~\cite{ALICE-TPC}. While implementations of fast simulations of given
sub-detectors is specific to their technology and experimental
environment and as such the responsibility of each LHC experiment, an
area of potential common investigation is frameworks for fast tuning and
validation.

The use of fully parametric response of experimental setups at the level
of reconstructed objects (tracks and clusters) for rapid evaluation of
physic's reach is also exploited by the experiments with generic
frameworks for fast simulation of typical collider detectors being used~\cite{bib-Delphes, PGS4}. 
The advantage of such frameworks is to allow
simultaneous studies for different experiments as well as their
potential use by the theory community. Experiments specific fully
parameterized simulations providing reconstructed objects compatible
with the experiment's analysis frameworks for systematic verifications
have also been emerging~\cite{Lamarr} .

It is reasonable to expect that the full variety of simulation options,
from `smeared reconstructed quantities' to parameterisation for specific
detectors to the detailed Geant4, will need to be exploited by the
experiments for different tasks. Seamless ways of providing them in a
transparent integrated way in the experiments' simulation frameworks
should continue to be pursued with Geant4 supporting hooks for such
integrations.

\hypertarget{technical-challenges-software-and-computing-1}{%
\subsection{Technical Challenges (Software and
Computing)}\label{technical-challenges-software-and-computing-1}}

The hardware landscape has been always setting direction as far as
software evolution was concerned. The recent adaptation of the
simulation code to multithreading (MT) turned out to be relatively
straightforward, but still took several years to adopt and validate.
Geant4 can now run in the MT mode efficiently, distributing the events
between the threads and sharing resources like the geometry description
or physics data and thus reduce the memory footprint of a simulation
while maintaining its throughput performance. The efficient utilization
of Single Instruction Multiple Data (SIMD) architectures, on the other
hand turned out to be more challenging. Not only was a limited part of
the code `vectorizable' but also, as the already mentioned GeantV R\&D
has shown, overheads related to gathering the data for vector processing
presented a significant challenges (tail handling, work balancing across
threads, etc.) to efficiently (development time wise and run time wise)
profit from it~\cite{GeantV}.

>>>>

% Content is stored in separate TeX files for managability
\input{input_introduction}

\input{input_generators}

\input{input_simulation}
